% ══════════════════════════════════════════════════════════════════════════════
%  CHAPTER 3: COSMIC EVOLUTION
% ══════════════════════════════════════════════════════════════════════════════

\chapter{Cosmic Evolution (Lecture 5-7)}

% ──────────────────────────────────────────────────────────────────────────────
\section{Dynamical Equations}
\subsection{1st Friedmann Equation}
\begin{foundation}
    Recall the FLRW metric:
    \[
        g_{\mu\nu} = \begin{pmatrix}
            -1 & 0 & 0 & 0 \\
            0 & \frac{a^2}{1-kr^2} & 0 & 0 \\
            0 & 0 & a^2 r^2 & 0 \\
            0 & 0 & 0 & a^2 r^2 \sin^2\theta
        \end{pmatrix}
    \]
    where $a(t)$ is the scale factor, and $k$ is the curvature constant. \\

    By brunches of tedious calculations from Christoffel symbols, Riemann curvature tensor(If you really want to do the calculation, you can used the equations in the appendix B.2, trust me, it was not a good time), we can find out the Ricci tensor and Ricci scalar as below:
    \[
        \begin{cases}
            R_{00} = -3\frac{\ddot{a}}{a} \\
            R = \frac{6}{a^2}\left( a \ddot{a} + \dot{a}^2 + k \right)
        \end{cases}
    \] \\

    Now plugging above results into the Einstein's field equation, we can get the 1st Friedmann equation as below: \\
    \begin{align*}
        R_{00} - \frac{1}{2}g_{00}R &= 8\pi G T_{00} \\
        -3\frac{\ddot{a}}{a} + \frac{1}{2}\frac{6}{a^2}\left( a \ddot{a} + \dot{a}^2 + k \right) &= 8\pi G \rho \\
        -3\frac{\ddot{a}}{a} \frac{a}{a} + 3\frac{1}{a^2}\left( a \ddot{a} + \dot{a}^2 + k \right) &= 8\pi G \rho \\
        -3\frac{\ddot{a}a}{a^2} + 3\frac{1}{a^2}\left(a\ddot{a}\right) + 3 \frac{1}{a^2}\left( \dot{a}^2 + k \right) &= 8\pi G \rho \\
        3\frac{\dot{a}^2 + k}{a^2} &= 8\pi G \rho \\
        3 H^2 + 3\frac{k}{a^2} &= 8\pi G \rho \\
        H^2 &= \frac{8\pi G}{3}\rho - \frac{k}{a^2}
    \end{align*}
\end{foundation}

\bigskip \noindent
\begin{theory}
    The 1st Friedmann equation is written as below:
    \[
        H^2 = \frac{\dot{a}^2}{a^2} = \frac{8\pi G}{3}\rho - \frac{k}{a^2}
    \]
    where $H$ is the Hubble parameter, $\rho$ is the energy density, $k$ is the curvature constant\\

    This equation tells us how the expansion rate $H$ changes with the energy(mass) density $\rho$, curvature $k$
\end{theory}


% ──────────────────────────────────────────────────────────────────────────────
\clearpage
%\bigskip \noindent \bigskip \noindent \bigskip \noindent
\subsection{2nd Friedmann Equation}
\begin{foundation}
    Similarly, we can get the 2nd Friedmann equation by using the results of Ricci tensor and Ricci scalar, and finding the trace of the Einstein's field equation:
    \begin{align*}
        R_{\mu\nu} - \frac{1}{2}g_{\mu\nu}R &= 8\pi G T_{\mu\nu} \\
        R_{\mu\nu}g^{\mu\nu} - \frac{1}{2}g_{\mu\nu}g^{\mu\nu}R &= 8\pi G T_{\mu\nu}g^{\mu\nu} \\
        R - \frac{1}{2} R \delta^\mu_\mu &= 8\pi G T \\
        - R &= 8\pi G T \\
        - (\frac{6}{a^2}\left( a \ddot{a} + \dot{a}^2 + k \right) ) &= 8\pi G (- \rho + 3p) \\
        \frac{a \ddot{a}}{a^2} + \frac{\dot{a}^2 + k}{a^2} &= - \frac{8\pi G}{6}(- \rho + 3p) \\
        \frac{\ddot{a}}{a} + \frac{8\pi G}{3}\rho &= \frac{4\pi G}{3}(\rho) - \frac{4\pi G}{3}(3p) \\
        \frac{\ddot{a}}{a} &= -\frac{4\pi G}{3}(\rho + 3p)
    \end{align*}
\end{foundation}

\bigskip \noindent
\begin{theory}
    The 2nd Friedmann equation is written as below:
    \[
        \frac{\ddot{a}}{a} = -\frac{4\pi G}{3}(\rho + 3p) 
    \]
    where $p$ is the pressure. \\

    This equation tells us how the expansion acceleration $\ddot{a}$ changes with the energy(mass) density $\rho$ and pressure $p$.
\end{theory}

\bigskip \noindent
\begin{remark}
    The Friedmann equations assumed that the universe is homogeneous and isotropic, which means the energy density $\rho$ and pressure $p$ are the same everywhere in the large scale of universe.
\end{remark}


% ──────────────────────────────────────────────────────────────────────────────
\clearpage
%\bigskip \noindent \bigskip \noindent \bigskip \noindent
\subsection{Continuity Equation}
\begin{intuition}
    Now we have two equations with three unknowns: $a$, $\rho$, and $p$. Thus, we need the "3rd Friedmann equation" to solve for these variables, and we usually call it the continuity equation. 
\end{intuition}

\bigskip \noindent
\begin{foundation}
    We can first take the time derivative of the 1st Friedmann equation:
    \begin{align*}
        \frac{\dot{a}^2}{a^2} &= \frac{8\pi G}{3}\rho - \frac{k}{a^2} \\
        \dot{a}^2 &= \frac{8\pi G}{3}\rho a^2 - k \\
        2\dot{a}\ddot{a} &= \frac{8\pi G}{3}(\dot{\rho} a^2 + \rho 2 a \dot{a}) - 0 \\
        \dot{a} \ddot{a} \frac{1}{a \dot{a}} &= \frac{4\pi G}{3}(\dot{\rho} a^2 \frac{1}{a \dot{a}} + \rho 2 a \dot{a} \frac{1}{a \dot{a}}) \\
        \frac{\ddot{a}}{a} &= \frac{4\pi G}{3}(\dot{\rho} \frac{a}{\dot{a}} + 2\rho) \\
        - \frac{4\pi G}{3}(\rho + 3p) &= \frac{4\pi G}{3}(\dot{\rho} \frac{a}{\dot{a}} + 2\rho) \\
        \dot{\rho} \frac{a}{\dot{a}} + 2\rho &= -\rho - 3p \\
        \dot{\rho} \frac{a}{\dot{a}} &= -3\rho - 3p \\
        \dot{\rho} &= - 3\frac{\dot{a}}{a}(\rho + p)\\
    \end{align*}
\end{foundation}

\bigskip \noindent
\begin{theory}
    The continuity equation is written as below:
    \[
        \dot{\rho} + 3H(\rho + p) = 0
    \]

    It tell us how the energy(mass) density $\rho$ changes with the size of the universe $a$.
\end{theory}

\clearpage
%\bigskip \noindent
\begin{example}{Another proof of the continuity equation}
    Assuming that the universe is Adiabatic Expansion (i.e. no heat exchange), can you derive the continuity equation by using the first law of thermodynamics?
    \[
        dE + pdV = dQ = 0
    \]

    \bigskip\textbf{Solution.}\quad \\
    \begin{align*}
        dE + pdV &= 0 \\
        d(\rho V) + pdV &= 0 \\
        d(\rho a^3) + pd(a^3) &= 0 \\
        a^3 d\rho + 3a^2 \rho da + 3a^2 p da &= 0 \\
        a^3 d\rho + 3a^2 (\rho + p) da &= 0 \\
        d\rho + 3\frac{da}{a} (\rho + p) &= 0 \\
        \dot{\rho} + 3H(\rho + p) &= 0
    \end{align*}
\end{example}

\bigskip \noindent
\begin{extra}
    There is actually a third way to derive the continuity equation, by using the Local conservation of energy-momentum tensor:
    \[
        \nabla_\mu T^{\mu\nu} = 0
    \]
    In fact, the derivation from the local conservation of energy-momentum tensor suggest that the continuity equation is just a consequence of local energy-momentum conservation, which gives the physical meaning of the continuity equation. \\

    The proof is left as an exercise for the reader. \\
    (Hints: watch the Relativity 110f by Eigenchris, \cite{Eigenchris_Relativity_Course})
\end{extra}

% ──────────────────────────────────────────────────────────────────────────────
%\clearpage
\bigskip \noindent \bigskip \noindent \bigskip \noindent
\subsection{Critical Energy Density and Density Parameters}




% ──────────────────────────────────────────────────────────────────────────────
%\clearpage
\bigskip \noindent \bigskip \noindent \bigskip \noindent
\subsection{Equation of State(EoS) and Energy Density Evolution}



% =══════════════════════════════════════════════════════════════════════════════
%\clearpage
\bigskip \noindent \bigskip \noindent \bigskip \noindent
\section{Cosmological constant}
\subsection{Origin of Cosmological Constant}
\begin{intuition}
    Recall the 2nd Friedmann equation:
    \[
        \frac{\ddot{a}}{a} = -\frac{4\pi G}{3}(\rho + 3p) 
    \]
    Assume that the universe contains normal matter which $p \geq 0$ and $\rho > 0$, then we have $\ddot{a} < 0$. \\

    When Einstein get this result in 1917, he found the result is unacceptable for his belief of a static universe, which requires $\ddot{a} = \dot{a} = 0$. Thus, he introduced the cosmological constant $\Lambda$, in order to balance the gravity of matter, but he cant give out a physical meaning of $\Lambda$ at that time. \\
\end{intuition}


\bigskip \noindent
\begin{foundation}
    After introducing the cosmological constant $\Lambda$, the Einstein's field equation becomes:
    \[
        R_{\mu\nu} - \frac{1}{2}Rg_{\mu\nu} + \Lambda g_{\mu\nu} = 8\pi G T_{\mu\nu}
    \]

    If we treat the extra term $\Lambda g_{\mu\nu}$ as an energy-momentum tensor, we can rewrite the equation as:
    \[
        R_{\mu\nu} - \frac{1}{2}Rg_{\mu\nu} = 8\pi G (T_{\mu\nu} + T^{(\Lambda)}_{\mu\nu})
    \]
    where $T^{(\Lambda)}_{\mu\nu} = -\frac{\Lambda}{8\pi G} g_{\mu\nu}$ is the energy-momentum tensor of the cosmological constant. \\
\end{foundation}

\bigskip \noindent
\begin{intuition}
    In a physic intuition, the extra term come from cosmological constant can be treated as a perfect fluid with energy density $\rho_\Lambda = \frac{\Lambda}{8\pi G}$ and pressure $p_\Lambda = -\rho_\Lambda$. Thus, the equation of LHS and RHS of the Einstein's field equation are still balanced, we just add an extra contribution of energy apart from the normal matter. \\
\end{intuition}

%\clearpage
%\bigskip \noindent
\begin{theory}
    The cosmological constant in energy-momentum tensor form can be written as:
    \[
        T^{(\Lambda)}_{\mu\nu} = \begin{pmatrix}
            \rho_\Lambda & 0 & 0 & 0 \\
            0 & -\rho_\Lambda & 0 & 0 \\
            0 & 0 & -\rho_\Lambda & 0 \\
            0 & 0 & 0 & -\rho_\Lambda
        \end{pmatrix}
    \]
    whether:
    \[
        \begin{cases}
            \rho_\Lambda = \frac{\Lambda}{8\pi G} = \text{constant} \\
            p_\Lambda = -\rho_\Lambda
        \end{cases}
    \]

    And we can see that the energy is constant throughout spacetime with negative pressure, and we said that the cosmological constant is conceptually equivalent to vacuum energy. \\
\end{theory}


\bigskip \noindent
\begin{extra}
    When Hubble shows that universe is expanding in 1929, the cosmological constant is no longer needed to balance the gravity of matter, thus it is largely abandoned, and Einstein even call it the "biggest blunder" in his life. \\

    Dramatically, the cosmological constant is revived in 1998, when the observation of Type Ia supernovae shows that the universe is accelerating expanding, Astronomers need to introduce the extra term of energy back in EFE to explain the accelerated expansion, and the cosmological constant are revived with a different physical meaning, which is the dark energy.
\end{extra}

\begin{figure}[htbp]
    \centering
    \begin{tikzpicture}[x=1cm,y=1cm]
        % axis
        \draw[->, thick, DeepSpace] (0,0) -- (14.2,0)
            node[right, DeepSpace] {year};

        % event boxes
        \node[draw=DeepSpace, fill=GoldBG, rounded corners=2pt, align=left, font=\scriptsize, text width=3.8cm] (A) at (2.1,1.9)
        {\textbf{1917: Introduction of $\Lambda$}\\Einstein adds a positive \\ cosmological constant $\Lambda > 0$ to allow a static universe.};

        \node[draw=DeepSpace, fill=GrayBG, rounded corners=2pt, align=left, font=\scriptsize, text width=4.3cm] (B) at (7.0,-2.0)
        {\textbf{1929: Expansion discovered}\\Hubble's observation supports an expanding universe, so $\Lambda = 0$ and largely abandoned.};

        \node[draw=DeepSpace, fill=GreenBG, rounded corners=2pt, align=left, font=\scriptsize, text width=3.8cm] (C) at (11.9,1.9)
        {\textbf{1998: Revival of $\Lambda$}\\Type Ia supernovae reveal accelerated expansion, so $\Lambda$ returns in the standard $\Lambda$CDM model, with $\Lambda > 0$ interpreted as dark energy.};

        % timeline dots and labels
        \fill[DeepSpace] (2.1,0) circle (1.5pt);
        \fill[DeepSpace] (7.0,0) circle (1.5pt);
        \fill[DeepSpace] (11.9,0) circle (1.5pt);

        \node[font=\scriptsize, text=DeepSpace] at (2.1,-0.55) {1917};
        \node[font=\scriptsize, text=DeepSpace] at (7.0,0.55) {1929};
        \node[font=\scriptsize, text=DeepSpace] at (11.9,-0.55) {1998};

        % arrows along the timeline
        %\draw[->, thick, DeepSpace] (2.1,0) -- (7.0,0);
        \draw[->, thick, DeepSpace] (2.1,0) -- (11.9,0);

        % connectors
        \draw[thin, DeepSpace] (2.1,0.1) -- (A.south);
        \draw[thin, DeepSpace] (7.0,-0.1) -- (B.north);
        \draw[thin, DeepSpace] (11.9,0.1) -- (C.south);
    \end{tikzpicture}
    \caption{Three major changes in the historical role of the cosmological constant.}
    \label{fig:cosmological-constant-history}
\end{figure}

%\bigskip \noindent
% explain the difference between cosmological constant and vacuum energy
%\clearpage
\begin{importantnote}
    Although we said they are equivalent, but the cosmological constant and vaccum energy are not exactly the same thing. \\

    They act like the same way that fills the spacetime with constant energy density and negative pressure, but the cosmological constant based on the observation of expansion of universe, we measure the value of $\Lambda$ as around $10^{-52}\,\mathrm{m}^{-2}$ by using CMB and supernovae. \\
    
    However, the vacuum energy is based on the quantum field theory, which predicts a value of vacuum energy density as around $10^{120}$ times larger than the observed value of $\Lambda$. This huge discrepancy is called the \textbf{cosmological constant problem}, which is one of the biggest unsolved problems in physics.
\end{importantnote}



% ──────────────────────────────────────────────────────────────────────────────
\bigskip \noindent \bigskip \noindent \bigskip \noindent
\subsection{Friedmann Equations with Cosmological Constant}
\bigskip \noindent
\begin{theory}
    By considering the cosmological constant in EFE, and using the similar method shown in the previous sections, we can get the modified Friedmann equations as below, \\

    the 1st Friedmann equations becomes:
    \[
        H^2 = \frac{8\pi G}{3}\rho - \frac{k}{a^2} + \frac{\Lambda}{3}
    \]
    and the 2nd Friedmann equations becomes:
    \[
        \frac{\ddot{a}}{a} = -\frac{4\pi G}{3}(\rho + 3p) + \frac{\Lambda}{3}
    \]
    and the continuity equation is unchanged:
    \[
        \dot{\rho} + 3H(\rho + p) = 0
    \]
\end{theory}

% -────────────────────────────────────────────────────────────────────────────
%\bigskip \noindent \bigskip \noindent \bigskip \noindent
\clearpage
\subsection{Newtonian Gravity with Cosmological Constant}
\begin{foundation}
    Let $p = 0$ for non-relativistic matter. Recall 00-component of Trace-formed EFE:
    \begin{align*}
        R_{00} &= 8\pi G (T_{00} - \frac{1}{2}g_{00}T) + \Lambda g_{00} \\
        R_{00} &= 8\pi G (\rho - \frac{1}{2}(-1)(-\rho + 3p)) + \Lambda (-1) \\
        R_{00} &= 4\pi G (\rho + 3p) - \Lambda \\
        R_{00} &= 4\pi G \rho - \Lambda
    \end{align*}
    In the Weak-Field limit (Newtonian limit), we have $R_{00} \approx - \frac{1}{2}\nabla^2 h_{00}$, and $h_{00} = -2\Phi$, where $\Phi$ is the Newtonian potential, thus we can rewrite above equation as:
    \[
        \nabla^2 \Phi = 4\pi G \rho - \Lambda
    \] \\

    By using superposition principle, we can solve the above equation in mass term and cosmological constant term separately. Mass term:
    \begin{align*}
        \nabla^2 \Phi_M &= 4\pi G \rho \\
        \Phi_M(r) &= -\frac{GM}{r}
    \end{align*}
    Then assuming spherical coordinate, the cosmological constant term can be solved as:
    \begin{align*}
        \nabla^2 \Phi_\Lambda &= -\Lambda \\
        \frac{1}{r^2} \frac{d}{dr} \left( r^2 \frac{d\Phi_\Lambda}{dr} \right) &= -\Lambda \\
        r^2 \frac{d\Phi_\Lambda}{dr} &= -\frac{1}{3}\Lambda r^3 + C_1 \\
        \frac{d\Phi_\Lambda}{dr} &= -\frac{1}{3}\Lambda r + \frac{C_1}{r^2} \\
        \Phi_\Lambda(r) &= -\frac{1}{6}\Lambda r^2 - \frac{C_1}{r} + C_2
    \end{align*}
    By letting $C_1 = C_2 = 0$, we can get the solution of cosmological constant term as:
    \[
        \Phi_\Lambda(r) = -\frac{1}{6}\Lambda r^2
    \]
\end{foundation}

\clearpage
%\bigskip \noindent
\begin{theory}
    The introduction of cosmological constant also yields a modification to the Newtonian gravity, which can be written as:
    \[
        \Phi(r) = -\frac{GM}{r} - \frac{1}{6}\Lambda r^2
    \]
    where the first term is the usual Newtonian potential, and the second term is the contribution of cosmological constant, which is a repulsive potential that grows with distance. \\

    We can also hence get the modified Newtonian force as below:
    \[
        F(r) = -m \nabla^2 \Phi(r) = -\frac{GM m}{r^2} + \frac{1}{3}\Lambda r m
    \]
    where the first term is the usual Newtonian force, and the second term is the contribution of cosmological constant, which is a repulsive force that grows with distance. \\
\end{theory}

\bigskip \noindent %
\begin{example}{Effect of Cosmological Constant}
    Consider a Sun-Earth like system, \\

    (a). estimate the threshold of $R$ above which the cosmological replusion force will dominate over the gravitational attraction. \\
    (b). calculate the ratio of the cosmological force to the gravitational force at the distance of Pluto, which is around $40$ AU. \\

    \bigskip\textbf{Solution.}\quad \\
    (a). The threshold of $R$ can be found by setting the two forces equal:
    \[
        \frac{GM m}{R^2} = \frac{1}{3}\Lambda R m
    \]
    Solving for $R$, we get:
    \[
        R = \left( \frac{3GM}{\Lambda} \right)^{1/3}
    \]
    Plugging in the values of $G$, $M$, and $\Lambda$:
    \begin{align*}
        R &= \left( \frac{3 \cdot 6.67430 \times 10^{-11} \cdot 1.989 \times 10^{30}}{2.036 \times 10^{-35}} \right)^{1/3} \\
        &= 2.694 \times 10^{18} \mathrm{m} \\
    \end{align*}
    which is around $90$ parsecs, which is much larger than the size of solar system, thus the cosmological constant has negligible effect on the dynamics of solar system. \\

    (b). The ratio of the cosmological force to the gravitational force at the distance of Pluto is:
    \begin{align*}
       \frac{F_\Lambda}{F_G} &= \frac{\frac{1}{3}\Lambda r m}{\frac{GM m}{r^2}} \\
         &= \frac{1}{3} \frac{\Lambda r^3}{GM} \\
         &= \frac{1}{3} \frac{2.036 \times 10^{-35} \cdot (40 \times 1.496 \times 10^{11})^3}{6.67430 \times 10^{-11} \cdot 1.989 \times 10^{30}} \\
         &= 1.1 \times 10^{-17}
    \end{align*}
    which is an extremely small number, thus the cosmological constant has negligible effect within the solar system, even at the distance of Pluto. \\
\end{example}




% ──────────────────────────────────────────────────────────────────────────────
\bigskip \noindent \bigskip \noindent \bigskip \noindent
\section{Evolution of Scale Factor}
\subsection{EoS of Cosmologic Constant}
\begin{theory}
    The equation of state of cosmological constant is:
    \[
        w_\Lambda = \frac{p_\Lambda}{\rho_\Lambda} = -1 < -\frac{1}{3}
    \] \\

    Assume flat universe with $k=0$, and the universe is dominated by cosmological constant, then the 1st Friedmann equation becomes:
    \[
        H^2 = \frac{8\pi G}{3}\rho_\Lambda = \frac{\Lambda}{3}
    \]
    By solving this ODE, we can get the solution of scale factor as below:
    \[
        a(t) = a_0 e^{H (t - t_0)} = a_0 e^{\sqrt{\frac{\Lambda}{3}} (t - t_0)}
    \]
\end{theory}

% ──────────────────────────────────────────────────────────────────────────────
\bigskip \noindent \bigskip \noindent \bigskip \noindent
\subsection{Scale Factor Equation}
\begin{theory}
\end{theory}








% =══════════════════════════════════════════════════════════════════════════════
\bigskip \noindent \bigskip \noindent \bigskip \noindent
\section{SCDM and $\Lambda$CDM Models}
\begin{theory}
    The total energy density of the universe can be expressed as:
    \[
        \Omega (t) = \Omega_{d}(t) + \Omega_{r}(t) + \Omega_{\Lambda}(t) + \Omega_{K}(t)
    \]
    where $\Omega_{d}$ is the contribution of matter, which is the sum of dark matter and ordinary matter $ \Omega_{d} = \Omega_{dm} + \Omega_{om}$, $\Omega_{r}$ is the contribution of radiation, $\Omega_{\Lambda}$ is the contribution of cosmological constant, and $\Omega_{K}$ is the contribution of curvature. \\
\end{theory}

\bigskip \noindent
\begin{theory}
    In a Einstein-de Sitter universe, which is a flat universe with only matter and no cosmological constant, i.e.:
    \[
        \Omega_{d} = 1, \quad \Omega_{r} = 0, \quad \Omega_{\Lambda} = 0, \quad \Omega_{K} = 0
    \]
    , the total energy density is:
    \[
        \Omega = \Omega_{d} = 1
    \]
    The model are also known as the \textbf{Standard Cold Dark Matter (SCDM)} model, which is the simplest cosmological model, and it predicts a decelerating universe with $\ddot{a} < 0$. \\

\end{theory}

At the late 1990s, the SCDM model is challenged by the acceleration of universe, thus the \textbf{$\Lambda$CDM} model is proposed, which is a flat universe with matter and cosmological constant $\Omega_{d} + \Omega_{\Lambda} = 1$, with $\Omega_{\Lambda} > 0$.






% =══════════════════════════════════════════════════════════════════════════════
\bigskip \noindent \bigskip \noindent \bigskip \noindent
\section{Dark Energy and the Accelerating Universe}
\subsection{Observational Evidence for Dark Energy}

% ──────────────────────────────────────────────────────────────────────────────
\bigskip \noindent \bigskip \noindent \bigskip \noindent
\subsection{Dark Energy Models}







% =══════════════════════════════════════════════════════════════════════════════
\bigskip \noindent \bigskip \noindent \bigskip \noindent
\section{Fate of Universe}
% different combination explain
\begin{foundation}
\end{foundation}

% different combination result
\begin{theory}
\end{theory}

\bigskip \noindent
\begin{extra}
    So, what is the fate of our universe? \\

    Based in the current CMB measurements, we know that our universe is basically flat $k \approx 0$, and the energy density proportion is showed as below:
    \[
        \begin{cases}
            \Omega_\Lambda \approx 0.68 \\
            \Omega_{dm} \approx 0.27 \\
            \Omega_{om} \approx 0.05 \\
            \Omega_{r} \approx 10^{-4} \\
            \Omega_{K} \approx 0
        \end{cases}
    \]

    Thus, our universe is dominated by the cosmological constant, and the scale factor will grow exponentially in the future, which means the universe will expand forever and become colder and darker, which is called the "Big Freeze" scenario. \\
\end{extra}